
title: "Chapterthree_spatial"
author: "Daniel Ritchie"
date: "2025-03-25"
output: html_document


```{r}
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(terra)
library(raster) 
```


```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(terra)
library(raster)  # Needed for raster math operations

# Load sightings data, removing NA values
sightings <- read.csv("~/Desktop/phd/ARFN/SSD/occurance_records.csv") %>%
  filter(!is.na(longitude), !is.na(latitude))

# Convert sightings to sf object
sightings_sf <- st_as_sf(sightings, coords = c("longitude", "latitude"), crs = 4326)

# Study site coordinates
sites <- data.frame(
  name = c("Muloorina Station", "Sturt National Park", "Kinchega National Park", "Ngarkat Conservation Park"),
  lat = c(-29.14, -29.05, -32.32, -35.43),
  lon = c(137.54, 141.30, 142.17, 140.36)
)
site_CHAP1 <- data.frame(
  name = c("Muloorina Station", "Dhilba Guuranda–Innes National Park"),
  lat = c(-29.14, -35.2278),
  lon = c(137.54, 136.8947)
)



sites_sf <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326)
sites_sf_chap1 <- st_as_sf(site_CHAP1, coords = c("lon", "lat"), crs = 4326)

# Load Australia map
australia <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")
australia_states <- ne_states(country = "Australia", returnclass = "sf")

# Define study area bounding box
study_bbox <- st_bbox(sightings_sf)  # Get bounding box from study sites
```


```{r}
# --- Import the downloaded files --- #
options(timeout = 600)  # Increase timeout to 10 minutes
download.file("http://thredds.northwestknowledge.net:8080/thredds/fileServer/TERRACLIMATE_ALL/data/TerraClimate_tmin_2023.nc", 
              destfile = "tmin.nc", mode = "wb")
```


```{r}
## Load imported NC files

pet <- rast("TerraClimate_pet_2023.nc")
ppt <- rast("TerraClimate_ppt_2023.nc")
tmin <-rast("tmin_2023.nc")
tmax <-rast("tmax_2023.nc")

# --- Compute mean precipitation and evapotranspiration --- #
ppt_mean <- mean(ppt,  na.rm = TRUE)
pet_mean <- mean(pet,  na.rm = TRUE)
tmin_mean <- mean(tmin,  na.rm = TRUE)
tmax_mean <- mean(tmax,  na.rm = TRUE)

# --- Crop to study area --- #
study_extent <- extent(c(study_bbox$xmin, study_bbox$xmax, study_bbox$ymin, study_bbox$ymax))
ppt_mean <- crop(ppt_mean, study_extent)
pet_mean <- crop(pet_mean, study_extent)
tmin_mean <- crop(tmin_mean, study_extent)
tmax_mean <- crop(tmax_mean, study_extent)

# --- Calculate aridity index --- #
aridity_index <- ppt_mean/pet_mean

# --- Convert raster to dataframe for ggplot --- #
aridity_index_df <- as.data.frame(aridity_index, xy = TRUE)
colnames(aridity_index_df) <- c("longitude", "latitude", "aridity")


tmin_mean_df <- as.data.frame(tmin_mean, xy = TRUE)
colnames(tmin_mean_df) <- c("longitude", "latitude", "tmin")

ppt_mean_df <- as.data.frame(ppt_mean, xy = TRUE)
colnames(ppt_mean_df) <- c("longitude", "latitude", "ppt")



tmax_mean_df <- as.data.frame(tmax_mean, xy = TRUE)
colnames(tmax_mean_df) <- c("longitude", "latitude", "tmax")

aridity_index_df <- aridity_index_df %>%
  mutate(category = case_when(
    is.infinite(aridity) ~ 'Humid',
    aridity >= 0.65 ~ 'Humid',
    aridity >= 0.5 & aridity < 0.65 ~ 'Dry sub-humid',
    aridity >= 0.2 & aridity < 0.5 ~ 'Semi-arid',
    aridity >= 0.05 & aridity < 0.2 ~ 'Arid',
    aridity < 0.05 ~ 'Hyper-arid'
  )) %>%
  mutate(category = factor(category, levels = c( 
                                                'Hyper-arid', 'Arid', 'Semi-arid', 
                                                'Dry sub-humid', 'Humid'), ordered = TRUE))


library(sf)
drainage_basins <- st_read("~/Desktop/phd/ARFN/SSD/42343_shp/rbasin_polygon.shp")
drainage_basins <- st_transform(drainage_basins, crs = st_crs(australia))
```




```{r}
# --- Plot study area with aridity data --- #
Study_area_chap3 <- ggplot() +
  geom_raster(data = aridity_index_df, aes(x = longitude, y = latitude, fill = category)) +
scale_fill_manual(values = c(
  'Hyper-arid' = '#d73027', # Orange-red
  'Arid' = '#fc8d59',  
  'Semi-arid' = '#fee08b',  
  'Dry sub-humid' = '#d9ef8b',  
  'Humid' = '#1a9850'),
  name = "Aridity Index"
) +
#geom_sf(data = drainage_basins, fill = NA, color = "white", size = 0.2) +
  #geom_sf(data = australia, fill = NA, color = "black", size=1) +
  geom_sf(data = australia_states, fill = NA, color = "black", size=1) +
  geom_sf(data = sightings_sf, color = "blue", alpha = 0.1, size = 0.5) +
  geom_sf(data = sites_sf, color = "black", size = 2.5) +
  geom_text(data = sites, aes(x = lon, y = lat, label = name),color = "black", hjust = -0.1, angle = 10, size = 4) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), ylim = c(study_bbox$ymin, study_bbox$ymax), expand = FALSE) +
  labs(title = "Study Sites with Aridity Index", x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(133, 149), ylim = c(-37, -25.5), expand = FALSE) +  # Zoom in
  theme_minimal()+
  theme(
    panel.grid = element_blank()
  )

Study_area_chap3
```
```{r}
# --- Load libraries ---
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)

# --- Load and clean ALA records ---
sightings <- read.csv("~/Desktop/phd/ARFN/SSD/occurance_records.csv") %>%
  filter(!is.na(longitude), !is.na(latitude))

# Convert to sf object
sightings_sf <- st_as_sf(sightings, coords = c("longitude", "latitude"), crs = 4326)

# --- Load Australia basemap ---
australia <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")
australia_states <- ne_states(country = "Australia", returnclass = "sf")

# --- Plot all ALA records across Australia ---
AUS_ALA_map <- ggplot() +
  geom_sf(data = australia, fill = "grey95", color = "grey50", size = 0.5) +
  geom_sf(data = australia_states, fill = NA, color = "grey40", size = 0.4) +
  geom_sf(data = sightings_sf, color = "#1f78b4", alpha = 0.3, size = 0.5) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  labs(
    title = "ALA Occurrence Records for Painted Dragons",
    x = "Longitude", y = "Latitude"
  ) +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

AUS_ALA_map

```

```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", AUS_ALA_map, width = 9, height = 6, dpi = 300)
```


```{r}
# --- Plot study area with aridity data --- #
Study_area_chap1 <- ggplot() +
  geom_raster(data = aridity_index_df, aes(x = longitude, y = latitude, fill = category)) +
scale_fill_manual(values = c(
  'Hyper-arid' = '#d73027', # Orange-red
  'Arid' = '#fc8d59',  
  'Semi-arid' = '#fee08b',  
  'Dry sub-humid' = '#d9ef8b',  
  'Humid' = '#1a9850'),
  name = "Aridity Index"
) +
#geom_sf(data = drainage_basins, fill = NA, color = "white", size = 0.2) +
  #geom_sf(data = australia, fill = NA, color = "black", size=1) +
  geom_sf(data = australia_states, fill = NA, color = "black", size=1) +
  geom_sf(data = sightings_sf, color = "blue", alpha = 0.1, size = 0.5) +
  geom_sf(data = sites_sf_chap1, color = "black", size = 2.5) +
 geom_text(data = site_CHAP1, aes(x = lon, y = lat, label = name),color = "black", hjust = -0.1, angle = 10, size = 4) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), ylim = c(study_bbox$ymin, study_bbox$ymax), expand = FALSE) +
  labs(title = "Study Sites with Aridity Index", x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(131, 147), ylim = c(-37, -26), expand = FALSE) +  # Zoom in
 theme_minimal()+
  theme(
    panel.grid = element_blank()
  )

Study_area_chap1
```
```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", Study_area_chap1, width = 8, height = 5.5, dpi = 300)
```


```{r}
tmin_mean_df$tmin_class <- cut(
  tmin_mean_df$tmin,
  breaks = c(-Inf, 0, 5, 10, 15, 20, Inf),
  labels = c("<0°C", "0-5°C", "5-10°C", "10-15°C", "15-20°C", ">20°C")
)
# --- Plot study area with aridity data --- #
Study_area_tmin <- ggplot() +
  geom_raster(data = tmin_mean_df, aes(x = longitude, y = latitude, fill = tmin_class)) +

    #scale_fill_viridis_c(
  #  option = "plasma",  # Try "viridis", "plasma", or "inferno"
 #   name = "Mean Min Temp (°C)"
#  ) +
  
  scale_fill_brewer(
    palette = "BuYlRd",  # Red-Yellow-Blue palette (reverse for cold=blue)
    name = "Temperature Class"
  ) +
#geom_sf(data = drainage_basins, fill = NA, color = "white", size = 0.2) +
  #geom_sf(data = australia, fill = NA, color = "black", size=1) +
  geom_sf(data = australia_states, fill = NA, color = "black", size=1) +
  geom_sf(data = sightings_sf, color = "blue", alpha = 0.2, size = 0.5) +
  geom_sf(data = sites_sf, color = "black", size = 2.5) +
  geom_text(data = sites, aes(x = lon, y = lat, label = name),color = "white", hjust = -0.1, angle = 10, size = 4) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), ylim = c(study_bbox$ymin, study_bbox$ymax), expand = FALSE) +
  labs(title = "Study Sites with Mean Min temp 2023", x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(133, 149), ylim = c(-37, -25.5), expand = FALSE) +  # Zoom in
  theme_minimal()

Study_area_tmin
```








```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", Study_area_tmin, width = 11.25, height = 6, dpi = 300)
```

```{r}


tmax_mean_df$tmax_class <- cut(
  tmax_mean_df$tmax,
  breaks = c(-Inf, 15, 20, 25, 30, 35, Inf),
  labels = c("<15°C", "15-20°C", "20-25°C", "25-30°C", "30-35°C", ">35°C")
)
# --- Plot study area with aridity data --- #
Study_area_tmax <- ggplot() +
  geom_raster(data = tmax_mean_df, aes(x = longitude, y = latitude, fill = tmax_class)) +

       #scale_fill_viridis_c(
  #  option = "plasma",  # Try "viridis", "plasma", or "inferno"
 #   name = "Mean Min Temp (°C)"
#  ) +
  
  scale_fill_brewer(
    palette = "BuYlRd",  # Red-Yellow-Blue palette (reverse for cold=blue)
    name = "Temperature Class"
  ) +
#geom_sf(data = drainage_basins, fill = NA, color = "white", size = 0.2) +
  #geom_sf(data = australia, fill = NA, color = "black", size=1) +
  geom_sf(data = australia_states, fill = NA, color = "black", size=1) +
  geom_sf(data = sightings_sf, color = "blue", alpha = 0.2, size = 0.5) +
  geom_sf(data = sites_sf, color = "black", size = 2.5) +
  geom_text(data = sites, aes(x = lon, y = lat, label = name),color = "white", hjust = -0.1, angle = 10, size = 4) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), ylim = c(study_bbox$ymin, study_bbox$ymax), expand = FALSE) +
  labs(title = "Study Sites with Mean Max temp 2023", x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(133, 149), ylim = c(-37, -25.5), expand = FALSE) +  # Zoom in
  theme_minimal()

Study_area_tmax
```

```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", Study_area_tmax, width = 11.25, height = 6, dpi = 300)
```

```{r}


ppt_mean_df$ppt_class <- cut(
  ppt_mean_df$ppt,
  breaks = c(-Inf, 10, 15, 20, 25, 30, 35, Inf),
  labels = c("<15mm","10-15mm", "15-20mm", "20-25mm", "25-30mm", "30-35mm", ">35mm")
)
# --- Plot study area with aridity data --- #
Study_area_ppt <- ggplot() +
  geom_raster(data = ppt_mean_df, aes(x = longitude, y = latitude, fill = ppt_class)) +

       #scale_fill_viridis_c(
  #  option = "plasma",  # Try "viridis", "plasma", or "inferno"
 #   name = "Mean Min Temp (°C)"
#  ) +
  
  scale_fill_brewer(
    palette = "BuYlRd",  # Red-Yellow-Blue palette (reverse for cold=blue)
    name = "Precipitation Class"
  ) +
#geom_sf(data = drainage_basins, fill = NA, color = "white", size = 0.2) +
  #geom_sf(data = australia, fill = NA, color = "black", size=1) +
  geom_sf(data = australia_states, fill = NA, color = "black", size=1) +
  geom_sf(data = sightings_sf, color = "blue", alpha = 0.2, size = 0.5) +
  geom_sf(data = sites_sf, color = "black", size = 2.5) +
  geom_text(data = sites, aes(x = lon, y = lat, label = name),color = "white", hjust = -0.1, angle = 10, size = 4) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), ylim = c(study_bbox$ymin, study_bbox$ymax), expand = FALSE) +
  labs(title = "Study Sites with Mean Max temp 2023", x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(133, 149), ylim = c(-37, -25.5), expand = FALSE) +  # Zoom in
  theme_minimal()

Study_area_ppt
```




```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", Study_area_ppt, width = 11.25, height = 6, dpi = 300)
```


```{r}

# Load required libraries
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(terra)
library(raster)
library(purrr)

# Define years range for averaging
years <- c(seq(from = 1995, to = 2015, by = 5), 2020, 2024)

# Load the previously downloaded ppt and pet files
ppt_files <- paste0('TerraClimate_ppt_', years, '.nc')  # Modify with your actual file paths if needed
pet_files <- paste0('TerraClimate_pet_', years, '.nc')

# Load raster stacks
ppt_rasters <- map(.x = ppt_files, ~ stack(.x))
pet_rasters <- map(.x = pet_files, ~ stack(.x))

# Crop the rasters to the study area
study_extent <- extent(c(study_bbox$xmin, study_bbox$xmax, study_bbox$ymin, study_bbox$ymax))
ppt_rasters <- map(.x = ppt_rasters, ~ crop(.x, study_extent))
pet_rasters <- map(.x = pet_rasters, ~ crop(.x, study_extent))

# Compute mean annual values for each raster stack over the years (1995 to 2024)
ppt_raster_avg <- calc(stack(ppt_rasters), fun = mean, na.rm = TRUE)
pet_raster_avg <- calc(stack(pet_rasters), fun = mean, na.rm = TRUE)

# Convert to data frames
ppt_df <- rasterToPoints(ppt_raster_avg) %>% as.data.frame() %>% rename(precipitation = layer)
pet_df <- rasterToPoints(pet_raster_avg) %>% as.data.frame() %>% rename(evapotranspiration = layer)

# Merge precipitation and evapotranspiration data
ai_df <- left_join(ppt_df, pet_df, by = c("x", "y"))

# Calculate the aridity index
ai_df <- ai_df %>% mutate(aridity_index = precipitation / evapotranspiration)

# Categorize aridity
ai_df <- ai_df %>% mutate(category = case_when(
  is.infinite(aridity_index) ~ 'Humid',
  aridity_index >= 0.65 ~ 'Humid',
  aridity_index >= 0.5 & aridity_index < 0.65 ~ 'Dry sub-humid',
  aridity_index >= 0.2 & aridity_index < 0.5 ~ 'Semi-arid',
  aridity_index >= 0.05 & aridity_index < 0.2 ~ 'Arid',
  aridity_index < 0.05 ~ 'Hyper-arid'
)) %>% mutate(category = factor(category, levels = c('Hyper-arid', 'Arid', 'Semi-arid', 'Dry sub-humid', 'Humid'), ordered = TRUE))

# Plot the final average aridity index
ggplot() +
  geom_raster(data = ai_df, aes(x = x, y = y, fill = category)) +
  geom_sf(data = australia_states, fill = NA, color = "black", size = 1) +
  scale_fill_manual(values = c('Hyper-arid' = '#d73027', 'Arid' = '#fc8d59', 'Semi-arid' = '#fee08b', 'Dry sub-humid' = '#d9ef8b', 'Humid' = '#1a9850'), guide = guide_legend(reverse = TRUE)) +
  theme_void(base_size = 18) +
  theme(legend.title = element_blank())

```



years <- c(seq(1995, 2015, by = 5), 2020, 2024)
```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(terra)
library(raster)  # Needed for raster math operations
library(purrr)

# Load sightings data, removing NA values
sightings <- read.csv("~/Desktop/phd/ARFN/SSD/occurance_records.csv") %>%
  filter(!is.na(longitude), !is.na(latitude))

# Convert sightings to sf object
sightings_sf <- st_as_sf(sightings, coords = c("longitude", "latitude"), crs = 4326)

# Study site coordinates
sites <- data.frame(
  name = c("Muloorina Station", "Sturt National Park", "Kinchega National Park", "Ngarkat Conservation Park"),
  lat = c(-29.14, -29.05, -32.32, -35.43),
  lon = c(137.54, 141.30, 142.17, 140.36)
)

sites_sf <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326)

# Load Australia map
australia <- ne_countries(country = "Australia", scale = "medium", returnclass = "sf")
australia_states <- ne_states(country = "Australia", returnclass = "sf")

# Define study area bounding box
study_bbox <- st_bbox(sightings_sf)  # Get bounding box from sightings

# Define study area extent for cropping
study_extent <- extent(c(study_bbox$xmin, study_bbox$xmax, study_bbox$ymin, study_bbox$ymax))

# Download and process TerraClimate data
years <- c(seq(from = 1995, to = 2024))

# Construct URLs and destination filenames for ppt and pet
download_data <- function(var, years) {
  lead <- paste0('http://thredds.northwestknowledge.net:8080/thredds/fileServer/TERRACLIMATE_ALL/data/TerraClimate_', var, '_')
  urls <- paste0(lead, years, '.nc')
  destfiles <- paste0('TerraClimate_', var, '_', years, '.nc')
  map2(.x = urls, .y = destfiles, ~ download.file(url = .x, destfile = .y, method = 'curl'))
  return(destfiles)
}
```


```{r}
ppt_files <- download_data('ppt', years)
pet_files <- download_data('pet', years)
```


```{r}
# Load raster stacks
ppt_rasters <- map(.x = ppt_files, ~ stack(.x))
pet_rasters <- map(.x = pet_files, ~ stack(.x))

# Crop to study area
ppt_rasters <- map(.x = ppt_rasters, ~ crop(.x, study_extent))
pet_rasters <- map(.x = pet_rasters, ~ crop(.x, study_extent))

# Compute mean annual values
ppt_rasters <- map(.x = ppt_rasters, ~ calc(.x, fun = mean, na.rm = TRUE))
pet_rasters <- map(.x = pet_rasters, ~ calc(.x, fun = mean, na.rm = TRUE))

# Convert raster to dataframe
ppt_df <- map(.x = ppt_rasters, ~ rasterToPoints(.x) %>% as.data.frame(.) %>% rename(precipitation = layer))
pet_df <- map(.x = pet_rasters, ~ rasterToPoints(.x) %>% as.data.frame(.) %>% rename(evapotranspiration = layer))

# Join data
ai_df <- map2(.x = ppt_df, .y = pet_df, ~ left_join(.x, .y))
ai_df <- map2(.x = ai_df, .y = years, ~ .x %>% mutate(year = as.character(.y)))
ai_df <- map_df(.x = ai_df, ~ bind_rows(.x))

# Calculate aridity index
ai_df <- ai_df %>% mutate(aridity_index = precipitation / evapotranspiration)

# Categorize aridity
ai_df <- ai_df %>% mutate(category = case_when(
  is.infinite(aridity_index) ~ 'Humid',
  aridity_index >= 0.65 ~ 'Humid',
  aridity_index >= 0.5 & aridity_index < 0.65 ~ 'Dry sub-humid',
  aridity_index >= 0.2 & aridity_index < 0.5 ~ 'Semi-arid',
  aridity_index >= 0.05 & aridity_index < 0.2 ~ 'Arid',
  aridity_index < 0.05 ~ 'Hyper-arid'
)) %>% mutate(category = factor(category, levels = c('Hyper-arid', 'Arid', 'Semi-arid', 'Dry sub-humid', 'Humid'), ordered = TRUE))
```


```{r}
# Plot aridity index with zoomed-in study area
ggplot() +
  geom_raster(data = ai_df, aes(x = x, y = y, fill = category)) +
  geom_sf(data = australia_states, fill = NA, color = "black", size = 1) +
  #geom_sf(data = sightings_sf, color = "blue", alpha = 0.2, size = 0.5) +
  
  scale_fill_manual(values = c('Hyper-arid' = '#d73027', 'Arid' = '#fc8d59', 'Semi-arid' = '#fee08b', 'Dry sub-humid' = '#d9ef8b', 'Humid' = '#1a9850'), guide = guide_legend(reverse = TRUE)) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), ylim = c(study_bbox$ymin, study_bbox$ymax)) +
  theme_void(base_size = 18) +
  facet_wrap(~year, ncol = 7) +
  theme(legend.title = element_blank())
```




```{r}
library(gganimate)

# Create the animation with year labels
aridity_animation <- ggplot(ai_df) +
  geom_raster(aes(x = x, y = y, fill = category)) +
  geom_sf(data = australia_states, fill = NA, color = "black", size = 1) +
  scale_fill_manual(values = c('Hyper-arid' = '#d73027', 
                               'Arid' = '#fc8d59', 
                               'Semi-arid' = '#fee08b', 
                               'Dry sub-humid' = '#d9ef8b', 
                               'Humid' = '#1a9850'), 
                    guide = guide_legend(reverse = TRUE)) +
  coord_sf(xlim = c(study_bbox$xmin, study_bbox$xmax), 
           ylim = c(study_bbox$ymin, study_bbox$ymax)) +
  theme_void(base_size = 18) +
  theme(legend.title = element_blank()) +
  # Add the year label
  geom_text(data = ai_df %>% distinct(year), aes(x = study_bbox$xmin + 0.1, y = study_bbox$ymax - 0.1, label = as.character(year)),
            color = "black", size = 6, fontface = "bold", hjust = 0) +
  geom_text(data = sites, aes(x = lon, y = lat, label = name),color = "black", hjust = -0.1, angle = 10, size = 4) +
  geom_sf(data = sites_sf, color = "black", size = 2.5) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         style = north_arrow_fancy_orienteering()) +
  coord_sf(xlim = c(133, 149), ylim = c(-37, -25.5), expand = FALSE) +  # Zoom in
  # Use transition_states to cycle through the 'year'
  transition_states(year, transition_length = 2, state_length = 1, wrap = TRUE) +
  ease_aes('linear')  # For a smooth transition between frames


```

```{r}
# Render the animation
animate(aridity_animation, nframes = 30, width = 800, height = 600, fps = 1)


```


```{r}
anim_save("aridity_animation.gif",animation =last_animation())
```




## Extract climate data and graph 

```{r}
# Load required libraries
library(ncdf4)      # For handling NetCDF files
library(terra)      # Raster manipulation
library(dplyr)      # Data wrangling
library(lubridate)  # Date handling
library(ggplot2)    # Plotting
library(tidyr)      # Data transformation

# --- Define Study Sites (Latitude & Longitude) ---
sites <- data.frame(
  name = c("Muloorina Station", "Sturt National Park", "Kinchega National Park", "Ngarkat Conservation Park"),
  lat = c(-29.14, -29.05, -32.32, -35.43),
  lon = c(137.54, 141.30, 142.17, 140.36)
)

# --- Function to Download & Extract Site-Specific NetCDF Data ---
download_and_extract <- function(varname, start_year = 1995, end_year = 2024) {
  base_url <- "http://thredds.northwestknowledge.net:8080/thredds/fileServer/TERRACLIMATE_ALL/data/"
  extracted_data <- data.frame()

  for (year in start_year:end_year) {
    nc_file <- paste0(varname, "_", year, ".nc")
    url <- paste0(base_url, "TerraClimate_", nc_file)
    
    if (!file.exists(nc_file)) {
      message(paste("Downloading:", nc_file))
      download.file(url, nc_file, mode = "wb")
    }

    # Open NetCDF file
    nc <- nc_open(nc_file)
    
    # Get time data
    time <- ncvar_get(nc, "time")
    start_date <- as.Date("1900-01-01") + time
    months <- format(start_date, "%Y-%m")  # Convert to Year-Month format

    # Extract variable data
    data_array <- ncvar_get(nc, varname)
    
    # Get latitude and longitude values
    lat_vals <- ncvar_get(nc, "lat")
    lon_vals <- ncvar_get(nc, "lon")

    # Convert to RasterBrick and extract values for study sites
    raster_stack <- rast(nc_file)
    site_data <- extract(raster_stack, sites[, c("lon", "lat")], method = "bilinear")

    # Create dataframe for each site
    for (i in 1:nrow(sites)) {
      site_df <- data.frame(date = months, value = site_data[, i + 1], site = sites$name[i])
      extracted_data <- rbind(extracted_data, site_df)
    }

    nc_close(nc)  # Close NetCDF file
  }
  
  return(extracted_data)
}
```


```{r}
library(ncdf4)
library(terra)
library(dplyr)
library(lubridate)
library(purrr)

# Define study site locations
sites <- data.frame(
  name = c("Muloorina Station", "Sturt National Park", "Kinchega National Park", "Ngarkat Conservation Park"),
  lat = c(-29.14, -29.05, -32.32, -35.43),
  lon = c(137.54, 141.30, 142.17, 140.36)
)

extract_from_nc <- function(varname, start_year = 1995, end_year = 2024) {
  # Pre-allocate list for results
  all_data <- list()
  
  for (year in start_year:end_year) {
    nc_file <- paste0(varname, "_", year, ".nc")
    
    if (!file.exists(nc_file)) {
      message(paste("Skipping missing file:", nc_file))
      next
    }
    
    # Open NetCDF file
    nc <- nc_open(nc_file)
    
    # Get proper time information
    time <- ncvar_get(nc, "time")
    time_units <- ncatt_get(nc, "time", "units")$value
    
    # Parse time units (handles different reference dates)
    if (grepl("days since", time_units)) {
      ref_date <- sub("days since ", "", time_units)
      dates <- as.Date(time, origin = as.Date(ref_date))
    } else {
      # Fallback to your original approach if units are different
      dates <- as.Date("1900-01-01") + time
    }
    
    # Convert to proper date format
    months <- format(dates, "%Y-%m")
    
    # Read data as SpatRaster
    raster_data <- rast(nc_file)
    
    # Extract all points at once
    site_values <- terra::extract(raster_data, sites[, c("lon", "lat")], method = "bilinear")
    
    # Create data frame for this year
    year_data <- map_dfr(1:nrow(sites), function(i) {
      data.frame(
        date = months,
        value = as.numeric(site_values[i, -1]), # -1 to exclude ID column
        site = sites$name[i],
        stringsAsFactors = FALSE
      )
    })
    
    all_data[[as.character(year)]] <- year_data
    nc_close(nc)
  }
  
  # Combine all years
  bind_rows(all_data)
}

# Parallel processing option (uncomment if you want to use it)
# library(furrr)
# plan(multisession) # Use future_map_dfr instead of map_dfr in the function

# Run extractions
tmax_data <- extract_from_nc("tmax")
tmin_data <- extract_from_nc("tmin")
ppt_data <- extract_from_nc("ppt")

# Save results
write.csv(tmax_data, "tmax_data.csv", row.names = FALSE)
write.csv(tmin_data, "tmin_data.csv", row.names = FALSE)
write.csv(ppt_data, "ppt_data.csv", row.names = FALSE)

```




```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Read the CSV files
tmax_data <- read.csv("tmax_data.csv")
tmin_data <- read.csv("tmin_data.csv")
ppt_data  <- read.csv("ppt_data.csv")

# Combine datasets
climate_data <- bind_rows(
  tmax_data %>% mutate(variable = "Tmax"),
  tmin_data %>% mutate(variable = "Tmin")
 # ppt_data  %>% mutate(variable = "Precipitation")
)


climate_data <- climate_data %>% separate(date, into =c("year", "month"), sep = "-", remove = FALSE)
climate_data$month <- as.numeric(climate_data$month)
climate_data$month <- factor(climate_data$month, levels =1:12, labels = month.abb, ordered = TRUE)
```


```{r}

# --- Compute Monthly Means & Standard Deviations for All Years ---
climate_summary <- climate_data %>%
  group_by(site, variable, month) %>%
  summarize(mean_value = mean(value, na.rm = TRUE),  # Average over all years
            sd_value = sd(value, na.rm = TRUE), .groups = "drop")
```





```{r}
temp_profile <- ggplot(
    climate_summary,
    aes(x = month, y = mean_value,
        color = site, linetype = variable, group = interaction(site, variable))) +
  geom_line(size = 0.5) +   
  geom_point(size = 1) +  
  geom_errorbar(aes(ymin = mean_value - sd_value,
                    ymax = mean_value + sd_value),
                width = 0.2) +  
  labs(
    x = "Month", y = expression(Temperature ~(degree *C)),
    color = "Site",
    linetype = "Climate Variable"
  ) +
  scale_x_discrete(limits = month.abb) +  
  scale_color_manual(values = c(
    "Muloorina Station" = "darkred",
    "Sturt National Park" = "#fc8d59",
    "Kinchega National Park" = "#4575b4",
    "Ngarkat Conservation Park" = "#91bfdb"
  )) + 
  theme_classic() + 
  theme(
    axis.title = element_text(size = 16, colour = "black"),   # black axis titles
    axis.text  = element_text(size = 14, colour = "black"),   # black tick labels
    axis.line  = element_line(linewidth = 1.2, colour = "black"),  # thicker axis lines
    legend.position = "none"
  )+
  facet_wrap(~site_group) 

temp_profile

```


```{r}
library(ncdf4)
library(terra)
library(dplyr)
library(lubridate)
library(purrr)

# ---- Study sites ----
sites <- data.frame(
  name = c("Muloorina Station", "Sturt National Park",
           "Kinchega National Park", "Ngarkat Conservation Park"),
  lat = c(-29.14, -29.05, -32.32, -35.43),
  lon = c(137.54, 141.30, 142.17, 140.36)
)

# ---- Extraction Function ----
extract_from_nc <- function(varname, start_year = 1995, end_year = 2024) {

  all_data <- list()

  for (year in start_year:end_year) {
    
    nc_file <- paste0(varname, "_", year, ".nc")
    if (!file.exists(nc_file)) {
      message("Skipping missing file: ", nc_file)
      next
    }

    # Open file
    nc <- nc_open(nc_file)

    # ---- Extract and interpret time ----
    time <- ncvar_get(nc, "time")
    units <- ncatt_get(nc, "time", "units")$value

    if (grepl("days since", units)) {
      origin <- gsub("days since ", "", units)
      dates <- as.Date(time, origin = origin)
    } else if (grepl("hours since", units)) {
      origin <- gsub("hours since ", "", units)
      dates <- as.POSIXct(time * 3600, origin = origin)
      dates <- as.Date(dates)
    } else {
      stop("Unknown time units: ", units)
    }

    # ---- Read raster and extract ----
    r <- rast(nc_file)
    vals <- terra::extract(r, sites[, c("lon", "lat")], method = "bilinear")

    # Remove ID column
    vals <- vals[, -1]

    # ---- Build tidy data frame ----
    df_year <- purrr::map_dfr(seq_len(nrow(sites)), function(i) {
      tibble(
        date = dates,
        value = as.numeric(vals[i, ]),
        site = sites$name[i]
      )
    })

    all_data[[as.character(year)]] <- df_year
    nc_close(nc)
  }

  bind_rows(all_data)
}

```

```{r}
tmax_data <- extract_from_nc("tmax")
tmin_data <- extract_from_nc("tmin")
ppt_data  <- extract_from_nc("ppt")

write.csv(tmax_data, "tmax_data.csv", row.names = FALSE)
write.csv(tmin_data, "tmin_data.csv", row.names = FALSE)

```



```{r}
library(ggridges)
library(dplyr)
library(lubridate)

# Define custom ordering
site_order <- c(
  "Muloorina Station",
  "Sturt National Park",
  "Kinchega National Park",
  "Ngarkat Conservation Park"
)

daily_tmax <- tmax_data %>%
  mutate(
    month = month(date),
    year  = year(date),
    site  = factor(site, levels = site_order)
  )

ggplot(daily_tmax,
       aes(x = value, y = site, fill = site)) +
  geom_density_ridges(alpha = 0.7, scale = 1.9) +
  labs(x = "Daily Tmax (°C)", y = "Population") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic()

```

```{r}
daily_tmax <- daily_tmax %>%
  mutate(clade = ifelse(site %in% site_order[1:2], "Northern", "Southern"))

ggplot(daily_tmax,
       aes(x = value, y = site, fill = site)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  facet_grid(clade ~ ., scales = "free_y", switch = "y") +
  theme_classic()

```


```{r}
library(fmsb)

seasonal <- climate_summary %>%
  pivot_wider(names_from = month, values_from = mean) %>%
  group_by(site) %>%
  summarize(across(everything(), mean))

radarchart(as.data.frame(seasonal))


```










```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", temp_profile , width = 11, height = 6, dpi = 300)
```



## Tub temperature data

```{r}
library(tidyverse)
library(lubridate)

path <- "~/Desktop/phd/Documents/Methods/ibutton_csv"

files <- list.files(path, full.names = TRUE, pattern = "\\.csv$")

ibutton_all <- map_df(files, function(f){

  fname <- basename(f)
  # remove .csv
  fname_clean <- str_remove(fname, "\\.csv")

  # split by underscore
  parts <- str_split(fname_clean, "_", simplify = TRUE)

  room       <- parts[1]   # C1, C2, etc
  position   <- parts[2]   # back, front
  treatment  <- parts[3]   # T1, T2, Warm, Cool, etc
  ID         <- parts[4]   # 293

  # read file (assumes two columns: Reading, Values)
  dat <- read_csv(f, col_types = cols()) %>%
    rename(Date = 1, Temperature = 2) %>%
    mutate(
      Date = dmy_hm(Date),
      Room = room,
      Position = position,
      Treatment = treatment,
      ID = ID
    )
})

ibutton_all <- ibutton_all %>%
  mutate(day = date(Date))


daily_treatment <- ibutton_all %>%
  group_by(Treatment, day) %>%
  summarise(
    daily_mean = mean(Temperature),
    daily_min = min(Temperature),
    daily_max = max(Temperature),
    .groups = "drop"
  )
overall_treatment <- daily_treatment %>%
  group_by(Treatment) %>%
  summarise(
    mean_temp = mean(daily_mean),
    mean_min = mean(daily_min),
    mean_max = mean(daily_max)
  )

```




```{r}
library(tidyverse)
library(lubridate)
ibutton_all <- ibutton_all %>%
  mutate(Treatment = recode(Treatment,
                            "T1" = "Warm",
                            "T2" = "Cool"))

# extract hour of day
ibutton_hourly <- ibutton_all %>%
  mutate(hour = hour(Date)) %>%
  group_by(Treatment, hour) %>%
  summarise(
    mean_temp = mean(Temperature),
    sd_temp   = sd(Temperature),
    n         = n(),
    se_temp   = sd_temp / sqrt(n),
    .groups = "drop"
  )
```


```{r}
treat_cols <- c(
  "Cool" = "#1f77b4",   # blue
  "Warm" = "#d62728"    # red
)

ggplot(ibutton_hourly,
       aes(x = hour, y = mean_temp, colour = Treatment)) +
  
  geom_line(size = 1) +
  geom_point() +
  
  geom_errorbar(
    aes(ymin = mean_temp - se_temp,
        ymax = mean_temp + se_temp),
    width = 0.25
  ) +
  
  scale_colour_manual(values = treat_cols) +
  
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  
  labs(
    x = "Hour of Day",
    y = "Temperature (°C)",
    colour = "Treatment"
  ) +
  theme_classic()

```

```{r}
## IMPORT  body temperature DATA

pictus_daily_bod <- read_csv("~/Desktop/phd/Documents/Chapter 3/Datasheets/pictus_bodytemp_2024.csv")
pictus_daily_bod$ID <- as.factor(pictus_daily_bod$ID)
pictus_daily_bod$Treatment <- as.factor(pictus_daily_bod$Treatment)

pictus_daily_bod <- pictus_daily_bod %>%
  mutate(
   Hour = hour(Start_Time) + minute(Start_Time)/60 + second(Start_Time)/3600
  )

calculate_exact_times <- function(data) {
  data %>%
    mutate(
      # Ensure times are in proper format
      Start_Time = as.POSIXct(Start_Time, format = "%H:%M"),
      End_time = as.POSIXct(End_time, format = "%H:%M")
    ) %>%
    group_by(Session, Date) %>%
    mutate(
      # Handle cases where start = end time
      session_duration = ifelse(
        Start_Time == End_time,
        60,  # assume 1 hour if no duration given
        as.numeric(difftime(End_time, Start_Time, units = "mins"))
      ),
      
      # Calculate order within session (0 to 1)
      session_min_order = min(Room_order, na.rm = TRUE),
      session_max_order = max(Room_order, na.rm = TRUE),
      normalized_order = (Room_order - session_min_order) / 
                        max(1, (session_max_order - session_min_order)),
      
      # Calculate exact time
      Exact_Time = Start_Time + (normalized_order * session_duration * 60),
      Hour_exact = hour(Exact_Time) + minute(Exact_Time)/60,
      
      # Also create a simple sequential time variable
      Simple_Hour = hour(Start_Time) + minute(Start_Time)/60 + 
                   (normalized_order * (session_duration/60))
    ) %>%
    ungroup()
}

# Apply the function
pictus_daily_bod <- calculate_exact_times(pictus_daily_bod)
```


```{r}

# Individual-level thermal experience
thermal_summary <- pictus_daily_bod %>%
  group_by(Treatment, ID) %>%
  summarise(
    ind_mean = mean(Body_temp, na.rm = TRUE),
    ind_min = min(Body_temp, na.rm = TRUE),
    ind_max = max(Body_temp, na.rm = TRUE),
    ind_range = ind_max - ind_min,
    .groups = 'drop'
  ) %>%
  group_by(Treatment) %>%
  summarise(
    n_individuals = n(),
    mean_daily_temp = round(mean(ind_mean), 1),
    mean_min_temp = round(mean(ind_min), 1),
    mean_max_temp = round(mean(ind_max), 1),
    mean_thermal_range = round(mean(ind_range), 1),
    sd_daily_temp = round(sd(ind_mean), 1),
    .groups = 'drop'
  )

print(thermal_summary)
```


```{r}
library(mgcv)

m <- gam(
  Body_temp ~ Treatment +
    s(Hour_exact, by = Treatment, k = 5) +
    s(ID, bs = "re"),
  data = pictus_daily_bod,
  method = "REML"
)
summary(m)
gam.check(m)

```

```{r}
# GAM model summary
gam_results <- summary(m)

cat("GAM MODEL SUMMARY - Treatment Effectiveness Evidence\n")
cat("===================================================\n")
cat("Deviance explained:", round(gam_results$dev.expl * 100, 1), "%\n")
cat("Adjusted R-squared:", round(gam_results$r.sq, 3), "\n")
cat("Sample size: n =", gam_results$n, "\n\n")

cat("SMOOTH TERMS - Pattern Differences:\n")
cat("Treatment 1 smooth (edf):", round(gam_results$s.table[1, "edf"], 2), "\n")
cat("Treatment 2 smooth (edf):", round(gam_results$s.table[2, "edf"], 2), "\n")
cat("→ Different complexities suggest distinct temporal patterns\n")
```




```{r}
# Create prediction data
new_data <- expand.grid(
  Hour_exact = seq(min(pictus_daily_bod$Hour_exact, na.rm = TRUE), 
             max(pictus_daily_bod$Hour_exact, na.rm = TRUE), 
             length = 100),
  Treat_Clade = unique(pictus_daily_bod$Treat_Clade),
  ID = pictus_daily_bod$ID[1]  # Reference individual
)

# Ensure factor levels match
new_data$Treat_Clade <- factor(new_data$Treat_Clade, 
                               levels = levels(pictus_daily_bod$Treat_Clade))
new_data$ID <- factor(new_data$ID, levels = levels(pictus_daily_bod$ID))

# Get predictions
pred <- predict(m_flexible, newdata = new_data, se.fit = TRUE, exclude = "s(ID)")

# Add predictions to new_data
new_data$fit <- pred$fit
new_data$se <- pred$se.fit
new_data$lower <- new_data$fit - 1.96 * new_data$se
new_data$upper <- new_data$fit + 1.96 * new_data$se

# Split the Treat_Clade back into separate columns for better plotting
new_data$Treatment <- sub("\\..*", "", new_data$Treat_Clade)
new_data$Clade <- sub(".*\\.", "", new_data$Treat_Clade)

# Do the same for original data for points
pictus_daily_bod$Treatment_plot <- sub("\\..*", "", pictus_daily_bod$Treat_Clade)
pictus_daily_bod$Clade_plot <- sub(".*\\.", "", pictus_daily_bod$Treat_Clade)
```


```{r}
# === CUSTOMIZE THESE VALUES ===
treatment_names <- c("1" = "Warm Treatment",  # Change "1" to your desired name
                     "2" = "Cool Treatment")   # Change "2" to your desired name

treatment_colors <- c("1" = "#9E2A2F",        # Change color for Treatment 1 (Sea Green)
                      "2" = "#4A90E2")        # Change color for Treatment 2 (Crimson)


plot_title <- "Body Temperature Patterns Throughout the Day"
plot_subtitle <- "Cool and Warm treatments produce distinct thermal trajectories"
y_axis_label <- "Body Temperature (°C)"
# === END CUSTOMIZATION ===
```


```{r}
# Create the plot with customizable elements
body_temp_graph <- ggplot(new_data, aes(x = Hour_exact, y = fit, 
                     color = Treatment, fill = Treatment)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  geom_point(data = pictus_daily_bod, 
             aes(x = Hour_exact, y = Body_temp, color = Treatment),
             alpha = 0.3, size = 1) +
  scale_color_manual(values = treatment_colors,
                     labels = treatment_names) +
  scale_fill_manual(values = treatment_colors,
                    labels = treatment_names) +
  labs(
    x = "Time of Day (Hours)",
    y = y_axis_label,
    color = "Treatment",
    fill = "Treatment"
  ) +
   facet_wrap(~ Clade, ncol = 2) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )
body_temp_graph
```
```{r}
ggsave("~/Desktop/phd/Documents/Chapter_1_Honours/TELO.tiff", body_temp_graph, width = 4, height = 3, dpi = 300)
```










```{r}
## IMPORT DATA
pictus_basking <- read_csv("~/Desktop/phd/Documents/Chapter 3/Datasheets/pictus_basking_data_2024.csv")
pictus_basking$ID <- as.factor(pictus_basking$ID)
pictus_basking$treatment <- as.factor(pictus_basking$treatment)
pictus_basking$BASKING_SCORE<- as.numeric(pictus_basking$BASKING_SCORE)
pictus_basking <- pictus_basking %>%
  mutate(
   Hour = hour(time_start) + minute(time_start)/60 + second(time_start)/3600
  )

```


```{r}
pictus_mean_basking <- pictus_basking %>%
  group_by(treatment, clade) %>%
  summarize(
    Mean_bask = mean(BASKING_SCORE  , na.rm = TRUE),
    SE_bask = sd(BASKING_SCORE  , na.rm = TRUE) / sqrt(sum(!is.na(BASKING_SCORE)))
  )
```


```{r}
library(mgcv)

mb <- gam(
  BASKING_SCORE ~ treatment +
    s(Hour, by = treatment, k = 10) +
    s(ID, bs = "re"),
  data = pictus_basking,
  method = "REML"
)
summary(mb)

```

```{r}
library(gratia)
library(ggplot2)

# Create prediction data
new_data <- with(pictus_basking, 
                 expand.grid(Hour = seq(min(Hour), max(Hour), length = 100),
                             treatment = unique(treatment),
                             ID = levels(ID)[1]))  # Use a reference ID

# Get predictions
pred <- predict(mb, newdata = new_data, se.fit = TRUE, exclude = "s(ID)")

# Add to new_data
new_data$fit <- pred$fit
new_data$se <- pred$se.fit
new_data$lower <- new_data$fit - 1.96 * new_data$se
new_data$upper <- new_data$fit + 1.96 * new_data$se

# Plot
ggplot(new_data, aes(x = Hour, y = fit, color = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  geom_jitter(data = pictus_basking, 
              aes(x = Hour, y = BASKING_SCORE, color = treatment),
              alpha = 0.3, size = 1, width = 0.1, height = 0) +
  labs(y = "Basking probability") +
  theme_minimal()
```




